Almost orthogonal vectors

### 2014-12-10 project3/paper/paper.tex
resctructured, added 3,6 and n,n+1'

\maketitle

@@

\section{$N = n+1$}

We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$. Before proving an upper bound on what the optimum value of $\epsilon$ can be, we look at the parallelogram law.

\subsection{Upper bound on minimum possible $\epsilon$}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.

@@

\begin{theorem}
If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the maximum value of $v_i\cdot v_j$ for $i\neq j$ is $\frac{-1}{n}$. This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin, and in this configuration $\epsilon=\frac{1}{n^2}$.  
\end{theorem}

\begin{proof}
Let $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, and let $\delta_{min}$ be the minimum possible $\delta$ obtained across all unit vectors $v_1, v_2, \ldots, v_{n+1}$. We first show that $\delta_{min} \le-\frac{1}{n}$ by showing that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\delta=-\frac{1}{n}$, and then will later show that $\delta_{min} \geq -\frac{1}{n}$ as well.

We must first describe the vectors that make up the origin centered simplex. To center our simplex at the origin we take each vector in the original simplex, and subtract from it the vector representing the centroid of the original simplex. The centroid of the original simplex takes the form $(\frac{1}{N},\frac{1}{N},\ldots,\frac{1}{N})$, so the points of the new simplex take the form
$$
\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we renormalize these vectors so that each vector has unit length, we get 
$$
\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
Now, taking the dot product between any two such vectors gives us
$$
\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
$$
So an upper bound on the value of $\delta_{min}$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\frac{-1}{n}$.

\section{Numerical analysis and associated conjectures}

@@

\section{Applications to sphere packing}
We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.

@@

>>>>

\maketitle
\tableofcontents

@@

\section{Some Optimal Configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{n=3,N=6}
In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are four possible dot products we get from these six vectors: $1,ab,-ab,a^2-b^2$. This gives us three possible values of $\epsilon$: $1,(a^2-b^2)^2,(ab)^2$. Let us consider the solution we get when $a^2-b^2=ab$. If we solve the resulting system according to the constraint $a^2+b^2=1$, we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we have that $\epsilon=\frac{1}{5}$, implying that these six points give us an optimal configuration.
\subsection{N=n+1}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we will find a configuration of vectors that achieves this lower bound. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.

@@

\begin{theorem}
If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{^2n}$. This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
\label{simplex}
\end{theorem}

\begin{proof}
Let $\epsilon$ be as before for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$. We show that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\epsilon=\frac{1}{n^2}$.

We must first describe the vectors that make up the origin centered simplex. To center our simplex at the origin we take each vector in the original simplex, and subtract from it the vector representing the centroid of the original simplex. The centroid of the original simplex takes the form $(\frac{1}{N},\frac{1}{N},\ldots,\frac{1}{N})$, so the points of the new simplex take the form
$$
\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we renormalize these vectors so that each vector has unit length, we get 
$$
\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
Now, taking the dot product between any two such vectors gives us
$$
\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
$$
So for this configuration $\epsilon=\frac{1}{n^2}$. By Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we have achieved the lower bound. 
\end{proof}
\section{Numerical analysis and associated conjectures}

@@

@@

\section{Variations}

We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$. Before proving an upper bound on what the optimum value of $\epsilon$ can be, we introduce the parallelogram law.

\subsection{Variations: Minimizing the dot product}
Let us consider the situation in which we try to minimize $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for $n+1$ vectors in an $n$ dimensional space.

\begin{theorem}
If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the minimum value of $v_i\cdot v_j$ for $i\neq j$ is $\frac{-1}{n}$. This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.  
\end{theorem}

\begin{proof}
Let $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, and let $\delta_{min}$ be the minimum possible $\delta$ obtained across all unit vectors $v_1, v_2, \ldots, v_{n+1}$. We computed in the proof of Theorem ~\ref{simplex} that for the origin centered simplex, we obtain a value of $\delta_{\min}=\frac{-1}{n}$, so $\delta_{\min}\le\frac{-1}{n}$. So we need only show that $\delta_{min} \geq -\frac{1}{n}$ as well.
